Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-11-12 Manuscript.Rmd
Updated the manuscript text and added mountain range names to the manuscript figure and to the maps in the app

# Correlation/fits for Initial CWD vs Delta CWD:
CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
CWD.lin.cor <- CWD.rels$lm1$r.squared^0.5 
CWD.dAIC <- round(CWD.rels$aic.cwd$AIC[1] - CWD.rels$aic.cwd$AIC[2], 0)

@@

There  are signs that tree species in western North America are in regional decline.  Johnston et al. (2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Peret et al. (2022) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2021); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerling 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

@@

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998) which appears to be a reasonable covariate for examining tree growth and mortality.  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration, PET); these variables are driven by temperature and precipitation, respectively.  CWD is geographically readily available from downscaled PRISM data (Daly et al. 2008, Wang et al. 2016).  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) found low-elevation California tree mortality best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 2001 the FIA program has gathered a variety of forestry data at plots along the west coast of the USA using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005).  The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used the revisit data to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.  They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.

The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (Climatic Water Deficit, initial CWD) and the change in water stress (delta CWD) over time.  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).  We constrained our analysis to only include remeasurements on areas considered forestland during both visits.

FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from FIA microplots.  Bechtold and Scott (2005) provide plot design details and measurement protocols.

Since 2000, FIA plots have been selected via spatially-balanced probability sampling (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.

Plots in the three states are sampled every 10 years with a target of 1/10th of plots sampled in a panel every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.  We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We found the 20-year mean of summer CWD values for each plot prior to and including the first visit year.  We refer this this value as the “initial CWD value”. We then found the mean of summer CWD values between visits and including the second visit year.  The difference between this second value and initial CWD value is the “delta CWD value”.  Both the initial CWD and delta CWD values for FIA plots are mapped in Figure 1.

## Analysis
Following Section 4.3.4 in Scott et al. (2005) we obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16).  The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.  The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.    The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.

The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth.  Mortality and growth values were standardized to a decadal time period.

The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).   The initial CWD values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).  The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated domain estimates for all CWD-based domains with 10 or more plots.

We examined contrasts in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.  The figure column "Above vs. Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.

@@

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above. The plot domain assignments for each species were sampled without replacement.  The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.  The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed in `r analysis.stats$strat.num` strata.

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).

Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots. Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC = $ `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).  Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).  Delta CWD values tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S4).

Individual species growth and mortality domain patterns were not always similar (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts. Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).  *Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.). Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).

Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`. There were fewer differences for "Above vs. Below Threshold" contrasts (`r mort.n.AvB`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.  *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3 spp.) or both Above and Below (3 spp.) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with 'significant' findings in both the Above and Below domains.  Five species had no significant contrasts.

The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).  Except where explicitly stated, we only discuss individual species' domain findings if the MC permutation p-values are < 0.001 (indicated by three asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)

@@

We hope our approach, results, and dashboard assist in refining future species inquiries.

>>>>

# Correlation/fits for Initial CWD vs Delta CWD:
CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
CWD.lin.cor <- format(round(CWD.rels$lm1$r.squared^0.5, 2), nsmall = 2) # Last digit = 0, want to show.
CWD.dAIC <- round(CWD.rels$aic.cwd$AIC[1] - CWD.rels$aic.cwd$AIC[2], 0)

@@

Numerous tree species in western North America exhibit increased mortality or decline.  van Mantgem et al. (2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010. Johnston et al. (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  Perret et al. (2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).  High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2022); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

@@

One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).  Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.  The Anderegg  et al. (2015) tree mortality study relied on CWD to capture drought occurrence.  Das et al. (2013) found low-elevation California tree mortality best modeled using CWD.  Restaino et al. (2016) found decreased Douglass fir growth with an increase in CWD.

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.  They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.  They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.

The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S. states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).  We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.  Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al. 2005).  We constrained our analysis to only include remeasurements for areas considered forestland during both visits.

FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measures, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.  Bechtold and Scott (2005) provide plot design details and measurement protocols.

The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.

Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.  We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year.  We then found the mean of summer CWD values between visits and including the second visit year.  We then defined 'delta CWD' values as the difference between this second value and initial CWD value.  Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.

## Analysis
We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al. 2005).  We expanded the number of sampled trees in each plot by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits. The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.  The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.

The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate. This produces a point estimate for a species’ mean growth.  We standardized mortality and growth values to a decadal time period.

We estimated, for each species examined, six domains based on CWD values (Figure 2C).   The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).  Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below). The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).  The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).  The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High. The Below domains are similarly labeled (BL, BM, BH). We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.  We limited the number of species examined to those with more than 400 occupied plots across the three states.  We calculated domain estimates for all CWD-based domains with 10 or more plots.

For each species we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) CWD domain estimates.  The figure column "Above vs. Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A). Negative estimates for the "Above vs. Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.  `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.

@@

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts. The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above. The plot domain assignments for each species were sampled without replacement.  Given this sampling approach, and since species distributions were not independent, the p-values are strictly interperetable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type). The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

@@

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews. Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` were gymnosperms.  The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years). Plots were distributed in `r analysis.stats$strat.num` strata.

Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.  Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.  Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).  Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).

Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots. Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC$ =  `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).  Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).  Delta CWD values tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S4).

Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B). Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains. All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts. Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values. *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.  Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).  *Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains. *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains. Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.). Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).

Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).  There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`). There were fewer differences for "Above vs. Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.  *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains. These results occurred in Above (3 spp.) or both Above and Below (3 spp.) domains. Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.  These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains.  Five species had no significant contrasts.

The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).  Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)

@@

Our analysis serves as an example of bivariate design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.  For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.  We believe our analysis approach can be refined for individual species analysis, with the potential for integrating a third variable that could introduce information besides a condition variable and change variable.  Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).  Our hope si that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
